### 2015-05-20 Harris_2015/manuscript-materials/draft.Rmd
ready to submit, maybe?

Ecologists' intense interest in drawing inferences about species interactions
---especially competition---from presence-absence data has a long history
[@macarthur_population_1958; @diamond_island_1975; @connor_checkered_2013]. If
nontrophic species interactions are important drivers of community assembly patterns,
then we should expect to see their influence in our data sets. Despite decades
of work and several major controversies, however [@lewin_santa_1983;
@strong_ecological_1984; @gotelli_swap_2003], existing methods for detecting
competition's effects on community structure are unreliable
[@gotelli_empirical_2009]. More generally, it can be difficult to reason about
the complex web of direct and indirect interactions that contribute to the
structure of real assemblages, especially when these interactions occur
against a background of other ecological processes such as dispersal and
environmental filtering [@connor_checkered_2013]. For this reason, it isn't
always clear what kinds of patterns would even constitute evidence of
competition, as opposed to some other biological process or random sampling
error [@roughgarden_competition_1983; @lewin_santa_1983].

Most existing methods in this field compare the frequency with which two
putative competitors are observed to co-occur against the frequency that would
be expected if *all* species on the landscape were independent
[@strong_ecological_1984; @gotelli_empirical_2009]. Examining a species pair
against such a "null" background, however, rules out the possibility that the
overall association between two species could be driven by an outside force,
as shown in Figure 1. Even though the two shrub species in this example compete 
with one another for resources at a mechanistic level, they end up clustering
together on the landscape because they both grow best in areas that are not overshadowed
by trees. Thus, even significant deviations from independence cannot---by 
themselves---tell us much about species' direct effects on one another.

@@

***Inferring $\alpha$ and $\beta$ coefficients from presence-absence data.***
The previous sections involved known values of $\alpha$ and $\beta$. In practice,
ecologists will often need to estimate these parameters from data instead. 
When the number of species is reasonably small, one can compute exact maximum 
likelihood estimates for all of the $\alpha$ and $\beta$ coefficients by optimizing 
Equation 1 (Appendix C).  Doing so allows us to find the unique set of $\alpha$ and
$\beta$ coefficients that would be expected to produce exactly the observed
co-occurrence frequencies. Fully-observed Markov
networks like the ones considered here have unimodal likelihood surfaces
[@murphy_machine_2012], ensuring that this procedure will always converge on
the global maximum.

For the analyses in this paper, I made a small modification to this maximum
likelihood procedure. Given the large number of parameters associated with
some of the networks to be estimated, I regularized the likelihood using a
logistic prior distribution [@gelman_weakly_2008] with a scale of 1 on the
$\alpha$ and $\beta$ terms.

The optimization procedures described here and in Appendix 2 are implemented
in the *rosalia* package^[source code available for download from
https://github.com/davharris/rosalia] for the R programming language
[@r_core_team_r_2015].

***Other inference techniques for comparison.***
After fitting Markov networks to the simulated landscapes described above, I
used several other techniques for inferring the sign and strength of marginal
associations between pairs of species.

@@

The third method, described in @gotelli_empirical_2009, involved simulating
possible landscapes from a null model that retained the row and column sums of
the original matrix [@strong_ecological_1984]. Using the default options in the
Pairs software described in @gotelli_empirical_2009, I simulated  the null
distribution of scaled C-scores (a test statistic describing the number of
*non*-co-occurrences between two species).  The software then calculated a
$Z$-statistic for each species pair using this null distribution. After
multiplying this statistic by $-1$ so that positive values corresponded to
facilitation and negative values corresponded to competition, I used it as
another estimate of species interactions.

@@

For the larger landscapes, I rescaled the four methods' estimates using linear
regression through the origin so that they all had a consistent
interpretation. In each case, I regressed the "true" $\beta$ coefficient for
each species pair against the model's estimate, re-weighting the pairs so that
each landscape contributed equally to the rescaled estimate^[The null model
generated one $Z$-score outlier greater than 1000, which dominated the
regression and squared error analyses. To reduce its influence on these
results, I changed its value to 32.5, which was the value of the next largest
$Z$-score in the null model's results.]. Then, for each combination of species
richness, landscape size, method, and presence/absence of environmental
filtering, I calculated the mean squared error associated with the rescaled
estimate across all 12 replicates. Finally, I calculated an $R^2$ value for
each landscape by comparing this squared error estimate with the squared error
associated with the null hypothesis that all interactions were zero.

@@

***Three species.***
As shown in Figure 1, the marginal relationship between the two shrub species
was positive---despite their competition for space at a mechanistic level---
due to indirect effects of the dominant tree species. As a result, the
covariance method falsely reported positive associations 94% of the time
and the randomization-based null model falsely reported such associations 100%
of the  time. The two methods for evaluating conditional relationships (Markov
networks and partial covariances), however, successfully controlled for the
indirect pathway via the tree species and each correctly identified the direct
negative interaction between the shrubs 92% of the time.

***Larger landscapes.***
The Markov network consistently outperformed all of the other techniques that
I evaluated (Figure 3). Of the remaining methods, partial covariances tended
to be most accurate, especially when 100 or more communities had been observed
on the landscape. $Z$-scores from the null model had the least predictive
utility overall, as indicated by $R^2$ values. In general, all the models'
accuracies declined when environmental filters were added, but the declines
were smaller when environmentally-induced correlations were spread out among a
larger number of pairwise interactions.

@@

The results presented above are very promising, as they show that Markov
networks can recover much of the variation in species' pairwise interaction
strengths from observational data, even when direct interactions are largely
overwhelmed by indirect effects (e.g. Figure 1) or environmental effects
(lower panels of Figure 3). For cases where it is infeasible to fit a Markov
network, these results also indicate that partial covariances---which can be
computed straightforwardly by linear regression---can often provide an
accurate approximation.

Apart from the environmental filters, the simulated landscapes presented here
represent the best-case scenario for these methods.  Future research should
thus examine these models' performance characteristics when the "true"
interaction matrices include guild structure or trophic levels, which could
make the $\beta$ coefficients much more difficult to infer (particularly for
linear approximations like the partial covariance approach;
@loh_structure_2013). On the other hand, ecologists may often have prior 
information about the nature of real species' interaction patterns from 
natural history or ecological experiments, which could substantially reduce
the probability and magnitude of error. The rosalia package has built-in
mechanisms for incorporating this kind of information, if it can be expressed
as a prior probability distribution or a penalty on the likelihood.

@@

Despite these limitations, the results with environmental filtering seem to
indicate that the method can be very robust.  Additionally, the fact that
Markov networks provide a likelihood function to optimize makes them highly
extensible, even when it is inconvenient to compute the likelihood exactly.
For example, the mistnet software package for joint species distribution modeling
[@harris_generating_2015] can fit *approximate* Markov networks to large
species assemblages (>100 species) while simultaneously modeling each species'
response to the abiotic environment with complex, nonlinear functions. This
sort of approach, which combines multiple ecological processes, could help
ecologists to disentangle different factors behind the co-occurrence patterns
we observe in nature. Numerous other extensions are possible: similar networks
can be fit with continuous variables, count data, or both
[@lee_learning_2012]. There are even methods [@whittam_species_1981;
@tjelmeland_markov_1998] that would allow the coefficient linking two species
in an interaction matrix to vary as a function of the abiotic environment or
of third-party species that could tip the balance between facilitation and
exploitation [@bruno_inclusion_2003]. Fully exploring these possibilities will
require more research into the various available approximations to the log-likelihood
and to its gradient, in order to balance efficiency, accuracy, and
the ability to generate confidence limits for statistical inference.

@@

This paper only scratches the surface of what Markov networks can do for
ecology. This family of models---particularly the Ising model for binary
networks---has been extremely well-studied in statistical physics for nearly a
century, and the models' properties, capabilities, and limits are well-understood
in a huge range of applications, from spatial lattices
[@gelfand_modelling_2005] to haphazard neural connections in the retina and
brain [@schneidman_weak_2006] to complex behavioral interactions among human
beings [@lee_statistical_2013]. Modeling species interactions using the same framework
would thus allow ecologists to tap into an enormous set of existing discoveries and
techniques for dealing with indirect effects, stability, and alternative
stable states.

Finally, the null models commonly used by ecologists for drawing inferences
about pairwise species interactions do not seem to distinguish between direct
and indirect interactions any better than simple correlation coefficients.
Null and neutral models can be very useful for clarifying our thinking about the numerical
consequences of species' richness and abundance patterns
[@harris_occupancy_2011; @xiao_strong_2015], but it is less clear whether deviations
from the null hypothesis can clearly implicate specific biological processes like 
competition.

@@

***Figure 3.***
Proportion of variance in interaction coefficients explained by each method
with 5, 10, or 20 species arrayed across varying numbers of sampled locations
when environmental filtering was absent (top row) or present (bottom  row).

>>>>

Ecologists' intense interest in drawing inferences about species interactions
---especially competition---from presence-absence data has a long history
[@macarthur_population_1958; @diamond_island_1975; @connor_checkered_2013]. If
nontrophic species interactions are important drivers of community assembly patterns,
then we should expect to see their influence in our data sets. Despite decades
of work and several major controversies, however [@lewin_santa_1983;
@strong_ecological_1984; @gotelli_swap_2003], existing methods for detecting
competition's effects on community structure are unreliable
[@gotelli_empirical_2009]. More generally, it can be difficult to reason about
the complex web of direct and indirect interactions in real assemblages,
especially when these interactions occur
against a background of other ecological processes such as dispersal and
environmental filtering [@connor_checkered_2013]. For this reason, it isn't
always clear what kinds of patterns would even constitute evidence of
competition, as opposed to some other biological process or random sampling
error [@roughgarden_competition_1983; @lewin_santa_1983].

Most existing methods in this field compare the frequency with which two
putative competitors are observed to co-occur against the frequency that would
be expected if *all* species on the landscape were independent
[@strong_ecological_1984; @gotelli_empirical_2009]. Examining a species pair
against such a "null" background, however, rules out the possibility that the
overall association between two species could be driven by an outside force.
For example, even though the two shrub species in Figure 1 compete 
with one another for resources at a mechanistic level, they end up clustering
together on the landscape because they both grow best in areas that are not overshadowed
by trees. If this sort of effect is common, then significant deviations 
from independence will not---by themselves---provide convincing evidence of
species' direct effects on one another.

@@

***Inferring $\alpha$ and $\beta$ coefficients from presence-absence data.***
The previous sections involved known values of $\alpha$ and $\beta$. In practice,
ecologists will often need to estimate these parameters from data instead. 
When the number of species is reasonably small, one can compute exact maximum 
likelihood estimates for all of the $\alpha$ and $\beta$ coefficients by optimizing 
Equation 1 (Appendix C).  Doing so allows us to find the unique set of $\alpha$ and
$\beta$ coefficients that would be expected to produce exactly the observed
co-occurrence frequencies. Fully-observed Markov
networks like the ones considered here have unimodal likelihood surfaces
[@murphy_machine_2012], ensuring that this procedure will always converge on
the global maximum. The optimization procedures described here and in Appendix C are 
implemented in the *rosalia* package [@harris_rosalia_2015] for the 
R programming language [@r_core_team_r_2015].

For analyses with 5 or more species, I made a small modification to the maximum
likelihood procedure described above. Given the large number of parameters 
associated with some of the networks to be estimated, I regularized the 
likelihood using a logistic prior distribution [@gelman_weakly_2008] with
a scale of 1 on the $\alpha$ and $\beta$ terms.

***Other inference techniques for comparison.***
After fitting Markov networks to the simulated landscapes described above, I
used several other techniques for inferring the sign and strength of marginal
associations between pairs of species (Appendix B).

@@

The third method, described in @gotelli_empirical_2009, involved simulating
possible landscapes from a null model that retains the row and column sums of
the original matrix [@strong_ecological_1984]. Using the default options in the
Pairs software described in @gotelli_empirical_2009, I simulated  the null
distribution of scaled C-scores (a test statistic describing the number of
*non*-co-occurrences between two species).  The software then calculated a
$Z$-statistic for each species pair using this null distribution. After
multiplying this statistic by $-1$ so that positive values corresponded to
facilitation and negative values corresponded to competition, I used it as
another estimate of species interactions.

@@

For the larger landscapes, I rescaled the four methods' estimates using linear
regression through the origin so that they all had a consistent
interpretation. In each case, I regressed the "true" $\beta$ coefficient for
each species pair against the model's estimate, re-weighting the pairs so that
each landscape contributed equally to the rescaled estimate^[The null model
generated one $Z$-score outlier greater than 1000, which dominated the
regression and squared error analyses. To reduce its influence on these
results, I changed its value to 32.5, which was the value of the next largest
$Z$-score in the null model's results.]. For each estimate of a species pair's
interactions, I used this regression to calculate the squared error associated
with method that produced it.  Finally, I averaged these squared errors for
each combination of species richness, landscape size, statistical method, and 
presence/absence of environmental filtering across all 12 replicates; the mean
squared errors associated with these subsets of the data determined the proportion
of variance explained by each method under different conditions.

@@

***Three species.***
As shown in Figure 1, the marginal relationship between the two shrub species
was positive---despite their competition for space at a mechanistic level---
due to indirect effects of the dominant tree species. As a result, the
covariance method falsely reported positive associations 94% of the time
and the randomization-based null model falsely reported such associations 100%
of the  time. The two methods for evaluating conditional relationships (Markov
networks and partial covariances), however, successfully controlled for the
indirect pathway via the tree species and each correctly identified the direct
negative interaction between the shrubs 94% of the time.

***Larger landscapes.***
The accuracy of the four evaluated methods varied substantially, depending
on the parameters that produced the simulated communities (Figure 3).
In general, however, there was a consistent ordering: the Markov network
explained 54% of the variance overall, followed by partial covariances (32%),
sample covariances (21%), and $Z$ scores from the null model (17%).  
The models' accuracies tended to decline when environmental filters were added, 
particularly when the number of species was small and the effects could not be 
diluted among many pairwise interactions.

@@

The results presented above are very promising, as they show that Markov
networks can recover species' pairwise interactions from observational  data,
even when direct interactions are largely overwhelmed by indirect effects
(e.g. Figure 1) or environmental effects (lower panels of Figure 3). For cases
where it is infeasible to fit a Markov network, these results also indicate
that partial covariances---which can be computed straightforwardly by linear
regression---can often provide an accurate approximation.

Apart from the environmental filters, the simulated landscapes presented here
represent the best-case scenario for these methods.  Future research should
thus examine these models' performance characteristics when the "true"
interaction matrices include guild structure or trophic levels, which could
make the $\beta$ coefficients much more difficult to infer (particularly for
linear approximations like the partial covariance approach;
@loh_structure_2013). On the other hand, ecologists may often have prior 
information about the nature of real species' interaction patterns from 
natural history or ecological experiments, which could substantially reduce
the probability and magnitude of error. The rosalia package [@harris_rosalia_2015]
has built-in mechanisms for incorporating this kind of information, if it can 
be expressed as a prior probability distribution or a penalty on the likelihood.

@@

Despite these limitations, the results with environmental filtering seem to
indicate that the method can be very robust.  Additionally, the fact that
Markov networks provide a likelihood function to optimize makes them highly
extensible, even when it is inconvenient to compute the likelihood exactly.
For example, the mistnet software package for joint species distribution modeling
[@harris_generating_2015] can fit *approximate* Markov networks to large
species assemblages (>300 species) while simultaneously modeling each species'
response to the abiotic environment with complex, nonlinear functions. This
sort of approach, which combines multiple ecological processes, could help
ecologists to disentangle different factors behind the co-occurrence patterns
we observe in nature. Numerous other extensions are possible: similar networks
can be fit with continuous variables, count data, or both
[@lee_learning_2012]. There are even methods [@whittam_species_1981;
@tjelmeland_markov_1998] that would allow the coefficient linking two species
in an interaction matrix to vary as a function of the abiotic environment or
of third-party species that could tip the balance between facilitation and
exploitation [@bruno_inclusion_2003]. Fully exploring these possibilities will
require more research into the various available approximations to the log-likelihood
and to its gradient, in order to balance efficiency, accuracy, and
the ability to generate confidence limits for statistical inference.

@@

This paper only scratches the surface of what Markov networks can do for
ecology. This family of models---particularly the Ising model for binary
networks---has been extremely well-studied in statistical physics for nearly a
century, and the models' properties, capabilities, and limits are well-
understood in a huge range of applications, from spatial modeling
[@gelfand_modelling_2005] to neuroscience [@schneidman_weak_2006] to models of
human behavior [@lee_statistical_2013; salakhutdinov_restricted_2007].
Modeling species interactions using the same framework would thus allow
ecologists to tap into an enormous set of existing discoveries and techniques
for dealing with indirect effects, stability, and alternative stable states.

These results also have important implications for the continued use of fixed-fixed
null models in ecology.  The small simulated landscapes described by
Figure 1 show that test statistics based on marginal co-occurrence (such as
C-scores) will not always have a straightforward relationship with the
underlying ecological processes.  Moreover, the larger communities analyzed in
Figure 3 often fell so far outside the null distribution^[Nearly 20% of the
species pairs fell outside the 99.99994% confidence intervals implied by their
$Z$ scores ($|Z| > 5$), and about 10% had uncorrected p-values below R's
default numerical precision of $2 \times 10^{-16}$.] that it probably makes 
more sense to reject whole model rather than to assign blame for the discrepancy to
any one species pair. On average, the pairwise $Z$ scores from the null
model provided less information about direct species interactions than correlation
coefficients did. Researchers using null modeling approaches may
be able to predict twice as much of the variance in species' "true" interaction
strengths using partial covariances from linear regression, or triple them 
using a Markov network.

Null and neutral models can be very useful for clarifying our thinking about
the numerical consequences of species' richness and abundance patterns
[@harris_occupancy_2011; @xiao_strong_2015], but deviations from a null model
must be interpreted with care [@roughgarden_competition_1983]. In complex
networks of ecological interactions---and even in small networks with three
species---it may simply not be possible to implicate individual species pairs
or specific ecological processes like competition by rejecting a general-
purpose null. Direct estimates of species' conditional associations may be the
only way to make these inferences reliably.

@@

***Figure 3.***
Proportion of variance in interaction coefficients explained by each method
with 5, 10, or 20 species arrayed across varying numbers of sampled locations
when environmental filtering was absent (top row) or present (bottom row). A
negative $R^2$ values implies that the squared error associated with 
the corresponding subset of the predictions was larger than the error one would
get from assuming that all coefficients equalled zero.
